Chemoinformatics: The Computer Science of Chemistry

### 2011-06-21 paper.tex
Fixed a variety of non ascii symbols

Both of these questions have relatively straightforward answers. We should care
about chemical information being made public since the drugs we take are the
result of intensive and costly research, and the more information is shared (a
process sometimes very hard to achieve in pharmaceutical companies) the more
easy it should be to develop novel treatments drugs. How chemical information
relates to computer science becomes clear when realizing the amount of data
available currently – 74 million different molecules with annotations are
currently stored in public databases such as PubChem with a large amount of
annotations, and the design of data structures for chemicals, as well as
subsequent data mining of this information, are areas where expertise in
computer science is dearly needed. (For a recent, detailed introduction to the
field written for computer scientists see \cite{brown2009} .) Cheminformatics comprises
different areas, which broadly speaking can be divided into three fields:
capturing data (using lab notebooks or potentially using formats such as the
Chemical Markup Language for publications); storing data (designing database
schemas, devising ontologies) and mining data (such as for bioactivity
prediction, which might involve algorithms based on graphs). However, what is a
major difference to the bioinformatics field is the type of information being
analyzed: While bioinformatics often deals with sequences, the domain of
cheminformatics is chemical structures. In case of the former, information can
very frequently be represented as one-dimensional strings, which are relatively
easy to handle computationally. On the other hand, chemical structures may
possess rings, branches as well as multiple valid representations of even the
same structure, potentially giving rise to ambiguities. Hence, chemical
structure is often thought of as being more difficult to standardize than
information from other domains. As an illustration, different representations of
the same chemical structure are shown in Figure 1. The top left hand corner
shows a structure a chemist would draw it – every ‘corner’ representing a carbon
atom (which can also be represented as a C), every O denoting oxygen, every N
nitrogen, every H hydrogen and so on. This is the standard format chemists use
when exchanging information with their peers. However, this is not directly
amenable to data mining -- different atoms might be collapsed into one node, sets
of atoms might have the same relevant properties assigned. Hence, the top right
hand corner of Figure 1 shows an annotated graph representation, taking into
account that atoms, though being of different elemental atom type, can have
similar properties -- e.g. both oxygen and nitrogen carry an 'On' label in this
graph.

@@

With the advent of high-throughput screening technologies, large libraries of
compounds can be screened against multiple targets in an efficient manner. Such
panel assays provide a broad, systems-level view of small molecule activities.
Models developed on such data afford us the opportunity to identify targets,
characterize off-target effects and so on. However, most approaches to this
problem are rather simplistic since they develop multiple individual models.
Instead one could imagine an approach that takes into account the correlations
in observed activities against multiple targets within a single model. Such an
approach should lead to more robust predictions for panel assays. Finally, does
this also better reflect clinical relevant compound profiles [SIDER database]
and a ‘personalized medicine’ concept (e.g. drug-drug interaction profiles) [The
Decision Tree].

>>>>

Both of these questions have relatively straightforward answers. We should care
about chemical information being made public since the drugs we take are the
result of intensive and costly research, and the more information is shared (a
process sometimes very hard to achieve in pharmaceutical companies) the more
easy it should be to develop novel treatments drugs. How chemical information
relates to computer science becomes clear when realizing the amount of data
available currently -- 74 million different molecules with annotations are
currently stored in public databases such as PubChem with a large amount of
annotations, and the design of data structures for chemicals, as well as
subsequent data mining of this information, are areas where expertise in
computer science is dearly needed. (For a recent, detailed introduction to the
field written for computer scientists see \cite{brown2009}). Cheminformatics
comprises different areas, which broadly speaking can be divided into three
fields: capturing data (using lab notebooks or potentially using formats such as
the Chemical Markup Language for publications); storing data (designing database
schemas, devising ontologies) and mining data (such as for bioactivity
prediction, which might involve algorithms based on graphs). However, what is a
major difference to the bioinformatics field is the type of information being
analyzed: While bioinformatics often deals with sequences, the domain of
cheminformatics is chemical structures. In case of the former, information can
very frequently be represented as one-dimensional strings, which are relatively
easy to handle computationally. On the other hand, chemical structures may
possess rings, branches as well as multiple valid representations of even the
same structure, potentially giving rise to ambiguities. Hence, chemical
structure is often thought of as being more difficult to standardize than
information from other domains. As an illustration, different representations of
the same chemical structure are shown in Figure 1. The top left hand corner
shows a structure a chemist would draw it -- every 'corner' representing a carbon
atom (which can also be represented as a C), every O denoting oxygen, every N
nitrogen, every H hydrogen and so on. This is the standard format chemists use
when exchanging information with their peers. However, this is not directly
amenable to data mining -- different atoms might be collapsed into one node,
sets of atoms might have the same relevant properties assigned. Hence, the top
right hand corner of Figure 1 shows an annotated graph representation, taking
into account that atoms, though being of different elemental atom type, can have
similar properties -- e.g. both oxygen and nitrogen carry an 'On' label in this
graph.

@@

With the advent of high-throughput screening technologies, large libraries of
compounds can be screened against multiple targets in an efficient manner. Such
panel assays provide a broad, systems-level view of small molecule activities.
Models developed on such data afford us the opportunity to identify targets,
characterize off-target effects and so on. However, most approaches to this
problem are rather simplistic since they develop multiple individual models.
Instead one could imagine an approach that takes into account the correlations
in observed activities against multiple targets within a single model. Such an
approach should lead to more robust predictions for panel assays. Finally, does
this also better reflect clinical relevant compound profiles [SIDER database]
and a 'personalized medicine' concept (e.g. drug-drug interaction profiles) [The
Decision Tree].
